---
title: "Group_4_IDS_Project"
Group: "Abhinav Bhandari, Megh Kurlekar, Sumit Kharche, Yash Singh"
output: html_document
date: "2025-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#install.packages("arrow")
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("data.table")
#install.packages("randomForest")
#install.packages("caret")
#install.packages("openxlsx")
#install.packages("corrplot")
#install.packages("ggcorrplot")
#install.packages("ggplot2")
#install.packages("tigris")
#install.packages("sf")
library(tidyverse)
library(arrow)
library(lubridate)
library(data.table)
#install.packages("randomForest")
library(randomForest)
library(caret)
library(openxlsx)
#install.packages("corrplot")
library(corrplot)
#install.packages("ggcorrplot")
library(ggcorrplot)
library(ggplot2)
install.packages("tigris")
library(tigris)
library(sf)
```


# 1. Data Reading & Initial Preparation

```{r}
# Load Static House Data
static_house_df = read_parquet("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/static_house_info.parquet")
```
```{r}
# Load July Weather Data for Each County
counties = unique(static_house_df$in.county)
weather_list = lapply(counties, function(cnty) {
  df = read_csv(paste0("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/weather/2023-weather-data/", cnty, ".csv"))
  df = df %>% 
    mutate(datetime = as.POSIXct(date_time)) %>% 
    filter(month(datetime) == 7) %>% 
    mutate(county = cnty)
})
weather_all = bind_rows(weather_list)
```

```{r}
# Load Energy Data for 10 Sample Houses for July
sample_housing = static_house_df$bldg_id[1:10]
energy_list = list()
for (id in sample_housing) {
  url = paste0("https://intro-datascience.s3.us-east-2.amazonaws.com/SC-data/2023-houseData/", id, ".parquet")
  df = read_parquet(url)
  df = df %>% mutate(datetime = as.POSIXct(time), bldg_id = id) %>%
    filter(month(datetime) == 7)
  energy_list[[as.character(id)]] = df
}
energy_usage_df = bind_rows(energy_list)
```
```{r}
# Merge Static, Energy, and Weather
static_house_df = static_house_df %>% rename(county = in.county)
combined_df = energy_usage_df %>%
  left_join(static_house_df, by = "bldg_id") %>%
  left_join(weather_all, by = c("datetime", "county")) %>%
  drop_na()
```


# 2. Data Cleaning
```{r}
all_zero_or_na = function(x) all(is.na(x) | x == 0)
cleaned_df = combined_df[, !sapply(combined_df, all_zero_or_na)]

if ("out.electricity.pv.energy_consumption" %in% names(cleaned_df)) {
  cleaned_df$out.electricity.pv.energy_consumption[cleaned_df$out.electricity.pv.energy_consumption < 0] = 0
}

cleaned_df = cleaned_df %>%
  mutate(out.total.energy = rowSums(across(contains(".energy_consumption")), na.rm = TRUE),
         in.cooling_setpoint = as.numeric(gsub("F", "", in.cooling_setpoint)),
         Day = day(datetime),
         Hour = hour(datetime))

```


# 4. Build Random Forest Model
```{r}
model_data = cleaned_df %>%
  select(
    energy = out.total.energy,
    temp = `Dry Bulb Temperature [°C]`,
    sqft = in.sqft,
    occupants = in.occupants,
    stories = in.geometry_stories,
    hour = Hour,
    county = county
  ) %>%
  mutate(county = as.factor(county)) %>%
  drop_na()

```

```{r}
set.seed(100)
split = createDataPartition(model_data$energy, p = 0.7, list = FALSE)
train = model_data[split, ]
test = model_data[-split, ]
```

```{r}
rf_model = randomForest(energy ~ ., data = train, ntree = 100)
# Save model to use in Shiny app
saveRDS(rf_model, file = "model.rds")
test$predicted = predict(rf_model, newdata = test)
```


# 5. Evaluate Model Accuracy
```{r}
rmse_value = RMSE(test$predicted, test$energy)
r2_value = cor(test$predicted, test$energy)^2
cat("Model RMSE:", round(rmse_value, 2), "\n")
cat("Model R²:", round(r2_value, 3), "\n")
```


# 6. Simulate +5°F Scenario
```{r}
temp_plus5 = test
temp_plus5$temp = temp_plus5$temp + 5
temp_plus5$predicted_plus5 = predict(rf_model, newdata = temp_plus5)
```


# 7. Peak Demand Visualizations
```{r}
# County-level Peak
county_peak_energy = temp_plus5 %>% group_by(county) %>% summarise(peak_kWh = max(predicted_plus5))

ggplot(county_peak_energy, aes(x = reorder(county, -peak_kWh), y = peak_kWh)) +
  geom_col(fill = "turquoise") +
  labs(title = "Peak Usage by County (+5°F)", y = "Peak Energy (kWh)") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1))
```


# Stories-level Peak
```{r}
story_peak_energy = temp_plus5 %>% group_by(stories) %>% summarise(peak_kWh = max(predicted_plus5))
```

```{r}
ggplot(story_peak_energy, aes(x = as.numeric(stories), y = peak_kWh)) +
  geom_line(color = "cyan", size = 1.2) +
  geom_point(size = 3, color = "red") +
  labs(title = "Trend for Peak Usage basaed on Number of Stories", x = "Stories", y = "Peak Energy (kWh)") +
  theme_minimal(base_size = 13)
```

# 8. Strategy to Reduce Peak
```{r}
temp_plus5$adjusted_usage = ifelse(temp_plus5$predicted_plus5 > quantile(temp_plus5$predicted_plus5, 0.9),
                                    temp_plus5$predicted_plus5 * 0.90, temp_plus5$predicted_plus5)
```

```{r}
adj_peak = max(temp_plus5$adjusted_usage)
peak_plus5 = max(temp_plus5$predicted_plus5)
reduction = peak_plus5 - adj_peak
```

```{r}
cat("Adjusted Peak After Strategy:", round(adj_peak, 2), "kWh\n")
cat("Peak Reduction:", round(reduction, 2), "kWh\n")
```

## Appliance Usage and Cooling System Plots

# 1. Energy Consumption by Appliance
```{r}
july_data = energy_usage_df %>% filter(month(datetime) == 7)
july_data_pivoted = july_data %>%
  select(datetime, contains(".energy_consumption")) %>%
  pivot_longer(cols = -datetime, names_to = "appliance", values_to = "consumption") %>%
  mutate(appliance = gsub("out\\.electricity\\.|\\.energy_consumption", "", appliance))
```

```{r}
# 1) Sum up consumption by appliance
july_summary = july_data_pivoted %>%
  group_by(appliance) %>%
  summarise(consumption = sum(consumption, na.rm = TRUE)) %>%
  ungroup()

# 2) Plot – modified from your screenshot
ggplot(july_summary, aes(x = fct_reorder(appliance, consumption), y = consumption, fill = consumption)) + geom_col(show.legend = FALSE) + coord_flip() + scale_fill_viridis_c(option = "C") + labs(title = "Energy Consumption by Appliance (July)", x = NULL, y = "Total Energy Consumption") + theme_minimal(base_size = 14) + theme(panel.grid.major.y = element_blank(), axis.text.y = element_text(size = 7)) + scale_fill_gradient(low = "cyan", high = "darkblue")

```
# 2. Total Electricity Usage by Cooling System
```{r}
df2 = cleaned_df %>% mutate(in.hvac_cooling_type = ifelse(in.hvac_cooling_type == 'None', 'No Cooling System', in.hvac_cooling_type))
```

```{r}
ggplot(df2, aes(x = in.hvac_cooling_type, y = out.total.energy, fill = in.hvac_cooling_type)) +
  geom_col() +
  labs(x = "Cooling System", y = "Total Energy (kWh)", title = "Total Energy Consumed Cooling System") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +   scale_fill_manual(values = c(
    "Central AC" = "navy",
    "No Cooling System" = "green"
  ))
  
```
# 3. Average Electricity Consumption per Cooling Setpoint
```{r}
average_setpoint = cleaned_df %>%
  group_by(in.cooling_setpoint) %>%
  summarise(avg_total_energy = mean(out.total.energy, na.rm = TRUE))
```

```{r}
ggplot(average_setpoint, aes(x = as.factor(in.cooling_setpoint), y = avg_total_energy)) +
  geom_bar(stat = "identity", fill = "orange") +
  labs(title = "Avg Electricity Consumption per Cooling Setpoint", x = "Cooling Setpoint", y = "Avg Energy (kWh)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))
```

```{r load_libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r}
# Energy Usage Peaks: hourly line chart
hourly_usage = cleaned_df %>%
  mutate(Hour = hour(datetime)) %>%
  group_by(Hour) %>%
  summarise(total_energy = sum(out.total.energy, na.rm = TRUE)) %>%
  ungroup()

ggplot(hourly_usage, aes(x = Hour, y = total_energy)) +
  geom_line(size = 1, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Hourly Energy Usage Peaks",
    x     = "Hour of Day",
    y     = "Total Energy (kWh)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#Key Contributors: pie chart of appliance energy consumption
appliance_contrib = cleaned_df %>%
  select(contains(".energy_consumption")) %>%
  pivot_longer(
    cols = everything(),
    names_to  = "component",
    values_to = "consumption"
  ) %>%
  group_by(component) %>%
  summarise(total = sum(consumption, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    component = gsub("out\\.electricity\\.|\\.energy_consumption", "", component),
    pct       = total / sum(total) * 100,
    label     = paste0(component, " (", round(pct, 1), "%)")
  )

ggplot(appliance_contrib, aes(x = "", y = total, fill = component)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(
    title = "Energy Consumption by Component"
  ) +
  theme_void() +
  theme(
    plot.title   = element_text(hjust = 0.5),
    legend.title = element_blank()
  )
```

```{r}
#Regional Variations: bar chart by county
regional_usage = cleaned_df %>%
  group_by(county) %>%
  summarise(total_energy = sum(out.total.energy, na.rm = TRUE)) %>%
  ungroup()

ggplot(regional_usage, aes(x = reorder(county, -total_energy), y = total_energy, fill = county)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Total Energy Usage by County",
    x     = "County",
    y     = "Total Energy (kWh)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title  = element_text(hjust = 0.5)
  )
```

## c) Modeling energy usage for July

```{r load-model-libs, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(caret)
#install.packages("gbm")
library(gbm)
```

```{r}
# Filter to July and pick your predictors
july_df = cleaned_df %>% 
  filter(month(datetime) == 7) %>% 
  mutate(
    Hour     = hour(datetime),
    Temp     = `Dry Bulb Temperature [°C]`,
    Setpoint = in.cooling_setpoint
  ) %>% 
  select(out.total.energy, Hour, Temp, Setpoint) %>% 
  drop_na()

# Split the dataset into 70% and 30% for training and testing respectively.
set.seed(42)
idx   = createDataPartition(july_df$out.total.energy, p = .7, list = FALSE)
train = july_df[idx, ]
test  = july_df[-idx, ]
```


```{r}
# Exploring various models
lm_mod    = lm(out.total.energy ~ Hour + Temp + Setpoint, data = train)
test$lm   = predict(lm_mod, test)

# Metrics
lm_rmse   = RMSE(test$lm, test$out.total.energy)
lm_r2     = R2(test$lm,  test$out.total.energy)
```

```{r}
gbm_mod   = gbm(
  formula      = out.total.energy ~ Hour + Temp + Setpoint,
  data         = train,
  distribution = "gaussian",
  n.trees      = 1000,
  interaction.depth = 3,
  shrinkage    = 0.01,
  cv.folds     = 5,
  verbose      = FALSE
)
best_iter = gbm.perf(gbm_mod, method = "cv")
test$gbm  = predict(gbm_mod, test, n.trees = best_iter)

gbm_rmse  = RMSE(test$gbm, test$out.total.energy)
gbm_r2    = R2(test$gbm, test$out.total.energy)
```


```{r}
poly_mod  = lm(out.total.energy ~ poly(Hour, 3) + Temp + Setpoint, data = train)
test$poly = predict(poly_mod, test)

poly_rmse = RMSE(test$poly, test$out.total.energy)
poly_r2   = R2(test$poly, test$out.total.energy)
```

```{r}
results = tibble(
  Model = c("Linear Reg.", "GBM", "Poly Reg."),
  RMSE  = c(lm_rmse, gbm_rmse, poly_rmse),
  R2    = c(lm_r2,   gbm_r2,   poly_r2)
)

knitr::kable(results, digits = 3, 
             caption = "July Energy Usage Models: RMSE & R²")
```



```{r future_scenarios_full, message=FALSE, warning=FALSE}
# 0. load libraries
library(dplyr)
library(lubridate)
library(caret)
install.packages("xgboost")
library(xgboost)
library(ggplot2)
library(tidyr)

# 1. Prepare July training data
july_df = cleaned_df %>%
  filter(month(datetime)==7) %>%
  mutate(
    Hour     = hour(datetime),
    Temp     = `Dry Bulb Temperature [°C]`,
    Setpoint = in.cooling_setpoint
  ) %>%
  select(out.total.energy, Hour, Temp, Setpoint) %>%
  drop_na()

set.seed(42)
split_idx = createDataPartition(july_df$out.total.energy, p=0.7, list=FALSE)
train_df   = july_df[ split_idx, ]
test_df    = july_df[-split_idx, ]

# 2. Train XGBoost (caret wrapper)
xgb_ctrl = trainControl(method="cv", number=5)
xgb_mod  = train(
  out.total.energy ~ Hour + Temp + Setpoint,
  data      = train_df,
  method    = "xgbTree",
  trControl = xgb_ctrl
)

# 3. Build prediction data for “current” and “future (current + 5°C)”
base_df = cleaned_df %>%
  filter(month(datetime)==7) %>%
  mutate(
    Hour = hour(datetime),
    Temp = `Dry Bulb Temperature [°C]`,
    Setpoint = in.cooling_setpoint
  ) %>%
  select(Hour, Temp, Setpoint)

orig_df = base_df
fut_df  = base_df %>% mutate(Temp = Temp + 5)

# 4. Predict outside of dplyr
pred_curr = predict(xgb_mod, newdata = orig_df)
pred_fut  = predict(xgb_mod, newdata = fut_df)

# 5. Combine & aggregate by hour
scenario_hourly = tibble(
  Hour          = orig_df$Hour,
  current_usage = pred_curr,
  future_usage  = pred_fut
) %>%
  group_by(Hour) %>%
  summarise(
    current_usage = sum(current_usage, na.rm=TRUE),
    future_usage  = sum(future_usage,  na.rm=TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols      = c(current_usage, future_usage),
    names_to  = "scenario",
    values_to = "energy_kWh"
  )

# 6. Plot
ggplot(scenario_hourly, aes(x=Hour, y=energy_kWh, fill=scenario)) +
  geom_col(position="dodge") +
  scale_x_continuous(breaks=0:23) +
  labs(
    title = "Current vs. Future (+5°C) Energy Usage by Hour",
    x     = "Hour",
    y     = "Energy (kWh)",
    fill  = "Scenario"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```